Data processing

### 2021-11-29 make_data.Rmd
Finish data processing script

---
title: "Make data files"
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

@@

Processing the files from `data-src` to produce data frames and igraph/network objects for analysis.

# Trades data

Load trades data from Zachary Richardson. Group trades into seasons using April 1 of each year.

```{r}
e <- new.env()
load(here::here("data-src", "data", "NBA_AnalysisData.RData"), envir=e)

trades <- e$trade.df_long |>
  select(-season) |> # This one is wrong
  mutate(
    season_from = case_when(
      month(Date) >= 4 ~ as.numeric(year(Date)),
      month(Date) <= 3 ~ as.numeric(year(Date)) - 1,
      TRUE ~ as.numeric(NA)
    ),
    season_to = season_from + 1,
    season = paste0(season_from, "-", season_to)
  )
```

@@

```{r graphlist}
graphlist <- lapply(
  sort(unique(trades$season)),
  function(s) {
    g <- graph_from_data_frame( 
      edb %>%
        filter(season == s) %>%
        select(P1, P2), 
      directed=FALSE)
    E(g)$weight <- count_multiple(g)
    g <- simplify(g)
    g$season <- s
    g
  }
)
```

Since that was built from edges, non-trading teams are inadvertently excluded. Here, we're adding them back in.

```{r graphlist-add-nodes}
graphlist[[3]] <- add.vertices(graphlist[[3]], 1, name="Spurs", label="Spurs", div="Central")

graphlist[[4]] <- add.vertices(graphlist[[4]], 1, name="Celtics", label="Celtics", div="Atlantic")
graphlist[[4]] <- add.vertices(graphlist[[4]], 1, name="Warriors", label="Warriors", div="Pacific")

graphlist[[5]] <- add.vertices(graphlist[[5]], 1, name="76ers", label="76ers", div="Atlantic")

graphlist[[9]] <- add.vertices(graphlist[[9]], 1, name="Nets", label="Nets", div="Atlantic")
graphlist[[9]] <- add.vertices(graphlist[[9]], 1, name="Suns", label="Suns", div="Pacific")

graphlist[[10]] <- add.vertices(graphlist[[10]], 1, name="Nuggets", label="Nuggets", div="Midwest")
graphlist[[10]] <- add.vertices(graphlist[[10]], 1, name="Knicks", label="Knicks", div="Atlantic")

graphlist[[11]] <- add.vertices(graphlist[[11]], 1, name="Suns", label="Suns", div="Pacific")
graphlist[[11]] <- add.vertices(graphlist[[11]], 1, name="Warriors", label="Warriors", div="Pacific")

graphlist[[12]] <- add.vertices(graphlist[[12]], 1, name="Mavericks", label="Mavericks", div="Midwest")

graphlist[[14]] <- add.vertices(graphlist[[14]], 1, name="Lakers", label="Lakers", div="Pacific")
graphlist[[14]] <- add.vertices(graphlist[[14]], 1, name="Wizards", label="Bullets", div="Atlantic")
graphlist[[14]] <- add.vertices(graphlist[[14]], 1, name="Jazz", label="Jazz", div="Midwest")

graphlist[[15]] <- add.vertices(graphlist[[15]], 1, name="Pacers", label="Pacers", div="Central")
graphlist[[15]] <- add.vertices(graphlist[[15]], 1, name="Celtics", label="Celtics", div="Atlantic")
graphlist[[15]] <- add.vertices(graphlist[[15]], 1, name="Knicks", label="Knicks", div="Atlantic")
graphlist[[15]] <- add.vertices(graphlist[[15]], 1, name="Cavaliers", label="Cavaliers", div="Central")

graphlist[[16]] <- add.vertices(graphlist[[16]], 1, name="76ers", label="76ers", div="Atlantic")
graphlist[[16]] <- add.vertices(graphlist[[16]], 1, name="Pacers", label="Pacers", div="Central")
graphlist[[16]] <- add.vertices(graphlist[[16]], 1, name="Nets", label="Nets", div="Atlantic")
graphlist[[16]] <- add.vertices(graphlist[[16]], 1, name="Magic", label="Magic", div="Atlantic")

graphlist[[18]] <- add.vertices(graphlist[[18]], 1, name="Heat", label="Heat", div="Atlantic")
graphlist[[18]] <- add.vertices(graphlist[[18]], 1, name="Celtics", label="Celtics", div="Atlantic")

graphlist[[19]] <- add.vertices(graphlist[[19]], 1, name="Kings", label="Kings", div="Pacific")
graphlist[[19]] <- add.vertices(graphlist[[19]], 1, name="Nets", label="Nets", div="Atlantic")
graphlist[[19]] <- add.vertices(graphlist[[19]], 1, name="Bulls", label="Bulls", div="Central")
graphlist[[19]] <- add.vertices(graphlist[[19]], 1, name="Cavaliers", label="Cavaliers", div="Central")

graphlist[[20]] <- add.vertices(graphlist[[20]], 1, name="Pacers", label="Pacers", div="Central")
graphlist[[20]] <- add.vertices(graphlist[[20]], 1, name="Rockets", label="Rockets", div="Midwest")
graphlist[[20]] <- add.vertices(graphlist[[20]], 1, name="Jazz", label="Jazz", div="Midwest")
graphlist[[20]] <- add.vertices(graphlist[[20]], 1, name="Mavericks", label="Mavericks", div="Midwest")

graphlist[[21]] <- add.vertices(graphlist[[21]], 1, name="76ers", label="76ers", div="Atlantic")
graphlist[[21]] <- add.vertices(graphlist[[21]], 1, name="Bulls", label="Bulls", div="Central")
graphlist[[21]] <- add.vertices(graphlist[[21]], 1, name="Cavaliers", label="Cavaliers", div="Central")
graphlist[[21]] <- add.vertices(graphlist[[21]], 1, name="Spurs", label="Spurs", div="Midwest")
graphlist[[21]] <- add.vertices(graphlist[[21]], 1, name="Clippers", label="Clippers", div="Pacific")

graphlist[[22]] <- add.vertices(graphlist[[22]], 1, name="Lakers", label="Lakers", div="Pacific")
graphlist[[22]] <- add.vertices(graphlist[[22]], 1, name="Jazz", label="Jazz", div="Midwest")
graphlist[[22]] <- add.vertices(graphlist[[22]], 1, name="Wizards", label="Wizards", div="Atlantic")
graphlist[[22]] <- add.vertices(graphlist[[22]], 1, name="Spurs", label="Spurs", div="Midwest")

graphlist[[23]] <- add.vertices(graphlist[[23]], 1, name="Heat", label="Heat", div="Atlantic")
graphlist[[23]] <- add.vertices(graphlist[[23]], 1, name="Pacers", label="Pacers", div="Central")
graphlist[[23]] <- add.vertices(graphlist[[23]], 1, name="Clippers", label="Clippers", div="Pacific")

graphlist[[24]] <- add.vertices(graphlist[[24]], 1, name="Heat", label="Heat", div="Atlantic")
graphlist[[24]] <- add.vertices(graphlist[[24]], 1, name="Jazz", label="Jazz", div="Midwest")
graphlist[[24]] <- add.vertices(graphlist[[24]], 1, name="Knicks", label="Knicks", div="Atlantic")

graphlist[[25]] <- add.vertices(graphlist[[25]], 1, name="Timberwolves", label="Timberwolves", div="Midwest")
graphlist[[25]] <- add.vertices(graphlist[[25]], 1, name="Nets", label="Nets", div="Atlantic")

graphlist[[26]] <- add.vertices(graphlist[[26]], 1, name="Jazz", label="Jazz", div="Midwest")

graphlist[[27]] <- add.vertices(graphlist[[27]], 1, name="Heat", label="Heat", div="Atlantic")
graphlist[[27]] <- add.vertices(graphlist[[27]], 1, name="Pacers", label="Pacers", div="Central")
graphlist[[27]] <- add.vertices(graphlist[[27]], 1, name="Timberwolves", label="Timberwolves", div="Midwest")
graphlist[[27]] <- add.vertices(graphlist[[27]], 1, name="Mavericks", label="Mavericks", div="Midwest")

graphlist[[28]] <- add.vertices(graphlist[[28]], 1, name="Heat", label="Heat", div="Atlantic")
graphlist[[28]] <- add.vertices(graphlist[[28]], 1, name="Lakers", label="Lakers", div="Pacific")

graphlist[[29]] <- add.vertices(graphlist[[29]], 1, name="Timberwolves", label="Timberwolves", div="Northwest")

graphlist[[30]] <- add.vertices(graphlist[[30]], 1, name="Mavericks", label="Mavericks", div="Southwest")
graphlist[[30]] <- add.vertices(graphlist[[30]], 1, name="Spurs", label="Spurs", div="Southwest")
graphlist[[30]] <- add.vertices(graphlist[[30]], 1, name="Warriors", label="Warriors", div="Pacific")

graphlist[[31]] <- add.vertices(graphlist[[31]], 1, name="Heat", label="Heat", div="Southeast")
graphlist[[31]] <- add.vertices(graphlist[[31]], 1, name="Clippers", label="Clippers", div="Pacific")
graphlist[[31]] <- add.vertices(graphlist[[31]], 1, name="Kings", label="Kings", div="Pacific")
graphlist[[31]] <- add.vertices(graphlist[[31]], 1, name="Wizards", label="Wizards", div="Southeast")
graphlist[[31]] <- add.vertices(graphlist[[31]], 1, name="Knicks", label="Knicks", div="Atlantic")

graphlist[[32]] <- add.vertices(graphlist[[32]], 1, name="Bucks", label="Bucks", div="Central")

graphlist[[33]] <- add.vertices(graphlist[[33]], 1, name="Hawks", label="Hawks", div="Southeast")

graphlist[[34]] <- add.vertices(graphlist[[34]], 1, name="Pacers", label="Pacers", div="Central")

graphlist[[35]] <- add.vertices(graphlist[[35]], 1, name="Pistons", label="Pistons", div="Central")

graphlist[[36]] <- add.vertices(graphlist[[36]], 1, name="Pistons", label="Pistons", div="Central")
graphlist[[36]] <- add.vertices(graphlist[[36]], 1, name="Suns", label="Suns", div="Pacific")

graphlist[[37]] <- add.vertices(graphlist[[37]], 1, name="Spurs", label="Spurs", div="Southwest")

graphlist[[38]] <- add.vertices(graphlist[[38]], 1, name="Magic", label="Magic", div="Southeast")

graphlist[[39]] <- add.vertices(graphlist[[39]], 1, name="Warriors", label="Warriors", div="Pacific")

graphlist[[41]] <- add.vertices(graphlist[[41]], 1, name="Timberwolves", label="Timberwolves", div="Northwest")

graphlist[[42]] <- add.vertices(graphlist[[42]], 1, name="Spurs", label="Spurs", div="Southwest")

graphlist[[43]] <- add.vertices(graphlist[[43]], 1, name="Warriors", label="Warriors", div="Pacific")

graphlist[[44]] <- add.vertices(graphlist[[44]], 1, name="Raptors", label="Raptors", div="Atlantic")
```

Save

```{r graphlist-save}
saveRDS(graphlist, file = "wgl.rds")
```

@@

@@

- `name` -- time-constant "franchise name"
- `label` -- time-varying team name
- `div` -- division name
- `season`, `season_from`, `season_to` -- season dates

```{r vdb}
vdb <- graphlist %>%
  setNames(nm = vapply(graphlist, graph_attr, character(1), name = "season")) %>%
  lapply(igraph::as_data_frame, what = "vertices") %>%
  bind_rows(.id = "season") %>%
  as_tibble() %>%
  separate(season, into = c("season_from", "season_to"), remove = FALSE) %>%
  mutate(
    across(starts_with("season_"), as.integer)
  )
vdb
```

@@

- `from`, `to` -- franchise names
- `season`, `season_from`, `season_to` -- season dates
- `weight` -- number of trades between `from` and `to` in `season`

```{r edb}
edb <- graphlist %>%
  setNames(nm = vapply(graphlist, graph_attr, character(1), name = "season")) %>%
  lapply(igraph::as_data_frame, what = "edges") %>%
  bind_rows(.id = "season") %>%
  as_tibble() %>%
  separate(season, into = c("season_from", "season_to"), remove = FALSE) %>%
  mutate(
    across(starts_with("season_"), as.integer),
    .from = pmin(from, to),
    .to = pmax(from, to)
  ) %>%
  select(
    from = .from, 
    to = .to, 
    season, 
    season_from,
    season_to,
    weight
  )
edb
```

@@

## Save

```{r save}
saveRDS(vdb, file = here::here("data", "df-vdb.rds"))
saveRDS(edb, file = here::here("data", "df-edb.rds"))
```

>>>>

---
title: "Data processing"
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

@@

Producing data frames and igraph objects for analysis.

# Team, franchises and divisions

Seasonal team and franchise names along with division memberships based on data on standings from https://www.basketball-reference.com.

```{r teams-franchises-divisions}
teams <- readRDS(here::here("data", "standings.rds")) %>%
  distinct(
    season_from,
    season_to,
    team = purrr::map_chr(strsplit(full_team_name, " "), tail, 1) |>
      recode("Orleans" = "Hornets"),
    division = gsub("Division", "", division) |> trimws()
  )

teams <- teams |>
  bind_rows(
    filter(teams, season_from == 2018) |>
      mutate(
        season_from = 2019,
        season_to = 2020
      )
  ) |>
  mutate(
    franchise = case_when(
      # Map season-teams to franchises
      season_from <= 1978 & team == "Braves" ~ "Clippers",
      season_from >= 2004 & season_from <= 2013 & team == "Bobcats" ~ "Hornets",
      team == "Bullets" & season_from <= 1996 ~ "Wizards",
      team == "Hornets" & season_from >= 2002 & season_from <= 2012 ~ "Pelicans",
      team == "SuperSonics" & season_from <= 2007 ~ "Thunder",
      TRUE ~ team
    ),
    season = paste0(season_from, "-", season_to)
  )
```

## Checking selected franchise histories

Franchises in columns, time in rows, team names in cells.

```{r selected-histories, R.options=list(knitr.kable.NA="")}
teams |>
  filter(franchise %in% c("Hornets", "Bobcats", "Pelicans", "Wizards", "Thunder", "Clippers")) |>
  select(team, franchise, season_from) |>
  tidyr::pivot_wider(names_from = franchise, values_from = team) |>
  arrange(season_from) |>
  knitr::kable()
```

# Data on trades

Process data on trades.

- Assign trades to seasons using April 1 as a cut off date.
- Columns with team names are called `franchise1` and `franchise2` as the names pretend to "season-invariant" names of the franchises, not actual team names used in during seasons.
- Fix some franchise names mislabeled in the original data.

```{r trades}
load(here::here("data-src", "data", "NBA_AnalysisData.RData"), 
     envir=NBA_AnalysisData <- new.env())

trades <- NBA_AnalysisData$trade.df_long |>
  as_tibble() |>
  select(-season) |>
  mutate(
    season_from = case_when(
      month(Date) >= 4 ~ as.numeric(year(Date)),
      month(Date) <= 3 ~ as.numeric(year(Date)) - 1,
      TRUE ~ as.numeric(NA)
    ),
    season_to = season_from + 1,
    season = paste0(season_from, "-", season_to)
  ) |>
  # Fix franchise/team names in trades data
  mutate(
    P1 = case_when(
      P1 == "Pelicans" & season_from >= 1988 & season_from <= 2001 ~ "Hornets",
      TRUE ~ P1
    ),
    P2 = case_when(
      P2 == "Pelicans" & season_from >= 1988 & season_from <= 2001 ~ "Hornets",
      TRUE ~ P2
    )
  ) |>
  mutate(
    franchise1 = pmin(P1, P2),
    franchise2 = pmax(P1, P2)
  ) |>
  select(-P1, -P2)

# Make sure of canonical order
stopifnot(with(trades, all(franchise1 < franchise2)))
```

Are there any franchise names in trade data that we can't match to standings on franchise name and season (expecting 0 rows in both)?

```{r}
trades %>%
  distinct(season_from, franchise1) %>%
  anti_join(teams, by = c("franchise1" = "franchise", "season_from")) %>%
  print(n = Inf)
```

```{r}
trades %>%
  distinct(season_from, franchise2) %>%
  anti_join(teams, by = c("franchise2" = "franchise", "season_from")) %>%
  print(n = Inf)
```

@@

```{r graphlist}
graphlist <- lapply(
  sort(unique(trades$season_from)),
  function(s) {
    # Select trades
    e <- trades %>%
      filter(season_from == s) %>%
      select(franchise1, franchise2)
    # Nodes
    n <- teams %>%
      filter(season_from == s) %>%
      select(franchise, team, division)
    g <- graph_from_data_frame(e, vertices = n, directed=FALSE)
    E(g)$weight <- count_multiple(g)
    g <- simplify(g, edge.attr.comb = list(weight = "min"))
    g$season <- paste0(s, "-", s+1)
    g
  }
)

saveRDS(graphlist, file = here::here("data", "igraph-list.rds"))
```

@@

Create node and edge data frames.

@@

- `name` -- season-invariant "franchise name"
- `team` -- team name
- `division` -- division name
- `season`, `season_from`, `season_to` -- season dates

```{r vdb}
vdb <- graphlist %>%
  setNames(nm = vapply(graphlist, graph_attr, character(1), name = "season")) %>%
  lapply(igraph::as_data_frame, what = "vertices") %>%
  bind_rows(.id = "season") %>%
  as_tibble() %>%
  separate(season, into = c("season_from", "season_to"), remove = FALSE) %>%
  mutate(
    across(starts_with("season_"), as.integer)
  )

vdb

saveRDS(vdb, file=here::here("data", "nodes.rds"))
```

@@

- `from`, `to` -- trading franchises' names
- `season`, `season_from`, `season_to` -- season dates
- `weight` -- number of trades between `from` and `to` in `season`

```{r edb}
edb <- graphlist %>%
  setNames(nm = vapply(graphlist, graph_attr, character(1), name = "season")) %>%
  lapply(igraph::as_data_frame, what = "edges") %>%
  bind_rows(.id = "season") %>%
  as_tibble() %>%
  separate(season, into = c("season_from", "season_to"), remove = FALSE) %>%
  mutate(
    across(starts_with("season_"), as.integer),
    .from = pmin(from, to),
    .to = pmax(from, to)
  ) %>%
  select(
    from = .from, 
    to = .to, 
    season, 
    season_from,
    season_to,
    weight
  )

edb

saveRDS(edb, file = here::here("data", "edges.rds"))
```

@@
